Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-03-07 Manuscript/manuscriptPS.tex
Updates in manuscript based on todo points

The molecular level interpretation of these observations is,
however, lacking, and  classical molecular dynamics (MD) simulations have been widely used in efforts to
to understand the PS headgroup structure, their influence on lipid bilayer properties, and their
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended strongly on the force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
the same force field without the NBfix parameters, combined with NMR chemical shifts and
REDOR \todo{Spell out REDOR?} experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

We have recently demonstrated that such controversies can be resolved by
comparing the C--H bond order parameters, $S_\mathrm{CH}$, of lipid headgroups between simulations
and experiments \cite{botan15,catte16}. The $S_\mathrm{CH}$ can be directly
measured from NMR experiments with high accuracy and compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}.
Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}.
%
Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
\todo{Should we leave the mention of ECC out from the Introduction
(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python program ({\tt scripts/calcOrderParameters.py}) that uses the
MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
The ion \todo{There are ion number densities, right?} number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
    the C--H bond order parameter values of -0.12 for the $\beta$-carbon, and +0.09 and -0.02
    for the $\alpha$-carbon.
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
  \todo{I think that the peak labeling would be good to show also in (A).}\\
%  \todo{Also for $\alpha$ these are OPs, right,  thus we could just exclude the word 'splittings'?}
%  \todo{Please confirm: Was the positive value used for $\alpha$ in the gray curve 0.02?}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS (T = 303 K, $^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  (T = 300 K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  \todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  We can do this just before submission when citation numbers will not change anymore.}
\end{figure}

@@

However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC forking of the $\alpha$-carbon \todo{observed in the experiments?} and the Slipids force field additionally correctly produces the significantly smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}) \todo{refer to experiments also?}.

%Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
%a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
%in the best three force field while other models give two maxima with different angles (Fig. \ref{dihedralsHG}).
Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
dihedral angle (single peak around 120$^{\circ}$) in comparison to the other models which yield a distribution between two angles (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and restricted\todo{rather: narrow?} angle distribution for PS than for PC headgroup in
CHARMM36 simulations that have the best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might reflect the increased rigidity anticipated\todo{detected, predicted, speculated?} in the early experimental studies \cite{browning80,buldt81}. %Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup. There is very little actual information in this statement. The variance was already noted above

@@

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. 
Counterion density profiles along membrane normal 
show significant differences between force fields in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in Gromos-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller areas (Fig. \ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values.
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36ua simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than with natrium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
In conclusion, the results are in line with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both counterion binding and hydrogen bonding network between lipid
headgroups \cite{petrache04}. \todo{can we really say this based on our data? We don't discuss hydrogen bonding}

@@

The experimentally observed modulation of headgroup order parameters
by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions and the lack of a ion-free reference state. The analysis is further complicated by the artificial aggregation of counterions
observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)\todo{Why isn't this a problem with PC}.
Therefore, we evaluate here the amount of bound charge not by adding salt
(although also this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
but by studying the changes of the headgroup order parameters with increasing amount of
negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.

>>>>

The molecular level interpretation of these observations is,
however, lacking, and  classical molecular dynamics (MD) simulations have been widely used in efforts to
to understand the PS headgroup structure, their influence on lipid bilayer properties, and their
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended strongly on the force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
the same force field without the NBfix parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

We have recently demonstrated that such controversies can be resolved by
comparing the C--H bond order parameters, $S_\mathrm{CH}$, of lipid headgroups between simulations
and experiments \cite{botan15,catte16}. The $S_\mathrm{CH}$ can be directly
measured from NMR experiments with high accuracy and compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}.
Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}.
%
%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
%\todo{Should we leave the mention of ECC out from the Introduction
%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python program ({\tt scripts/calcOrderParameters.py}) that uses the
MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
    the C--H bond order parameter values of -0.12 for the $\beta$-carbon, and +0.09 and -0.02
    for the $\alpha$-carbon.
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
  \todo{I think that the peak labeling would be good to show also in (A).
  There is possibly a new version of (C) coming.}\\
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS (T = 303 K, $^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  (T = 300 K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  \todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  We can do this just before submission when citation numbers will not change anymore.}
\end{figure}

@@

However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
also observed in the experiments (Fig. \ref{HGorderParameters}).

Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
dihedral angle (single peak around 120$^{\circ}$) in comparison to the other models which yield a distribution between two angles (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and narrow angle distribution for PS than for PC headgroup in
CHARMM36 simulations that have the best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
% Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup. There is very little actual information in this statement. The variance was already noted above

@@

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. 
Counterion density profiles along membrane normal 
show significant differences between force fields in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in Gromos-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller areas (Fig. \ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values.
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36ua simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
In conclusion, the results are in line with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.

@@

The experimentally observed modulation of headgroup order parameters
by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions and the lack of a ion-free reference state.
The analysis is further complicated by the artificial aggregation of counterions
observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)
% \todo{Why isn't this a problem with PC}.
% SAMULI: The difference between ion-free state (present in PC) and low amount of monovalent ions is larger than
% between counterion only and low amount of monovalent ions in charged lipids. I believe that therefore we can
% make conclusions from smaller ion concentrations without aggregation in PC lipids, but not in charged lipids.
Therefore, we evaluate here the amount of bound charge not by adding salt
(although also this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
but by studying the changes of the headgroup order parameters with increasing amount of
negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.

### 2019-03-07 Manuscript/manuscriptPSsuppl.tex
Updates in manuscript based on todo points

\subsection{Slipids}
{\it DOPC with 288 lipids.}
The starting structure for simulation was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}: 288 DOPS lipids, 11232 water molecules and 288 Na ions. The Slipids~\cite{jambeck12,jambeck2012another} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and Ions are described by the parameters derived by Aqvist~\cite{aqvist90}.
Simulation was performed for 200~ns, and the last 100~ns were used for the analysis. Simulation was carried out within the NPT ensemble using the GROMACS 5.0.4 package~\cite{abraham2015gromacs}. Timestep of 2~fs was used with the leapfrog integrator. The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath. Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
barostat~\cite{parrinello81} with a time constant of 10.0~ps. Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}. A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space. Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, whereas water molecules were constrained using SETTLE~\cite{miyamoto92}. Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.

@@

@@

\noindent
{\it DOPS and POPS bilayers.} \\
The data in Table \ref{PSsystems} and Refs. \citenum{slipidsPOPS298K,slipidsDOPS303K,slipidsDOPSfiles}.
\todo{T. Piggot and F. Favela, please write simulation details} \\

>>>>

\noindent
{\it DOPS with 288 lipids.}
The starting structure for simulation was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}: 288 DOPS lipids, 11232 water molecules and 288 Na ions. The Slipids~\cite{jambeck12,jambeck2012another} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and Ions are described by the parameters derived by Aqvist~\cite{aqvist90}.
Simulation was performed for 200~ns, and the last 100~ns were used for the analysis. Simulation was carried out within the NPT ensemble using the GROMACS 5.0.4 package~\cite{abraham2015gromacs}. Timestep of 2~fs was used with the leapfrog integrator. The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath. Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
barostat~\cite{parrinello81} with a time constant of 10.0~ps. Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}. A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space. Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, whereas water molecules were constrained using SETTLE~\cite{miyamoto92}. Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.

@@

%{\it DOPS and POPS bilayers.} \\
%The data in Table \ref{PSsystems} and Refs. \citenum{slipidsPOPS298K,slipidsDOPS303K,slipidsDOPSfiles}.

@@
